# Package Loading
```{r, warning = FALSE, message = FALSE}
library(readr)
```



```{r}
fetal_health <- read_csv("C:/Users/eacac/Documents/LUC/LUC-github/SP2024/STAT 437/Final Project/337-437-Final-Project/Data/fetal_health.csv", show_col_types = FALSE)
head(fetal_health, n = 10)
```

```{r}
colnames(fetal_health)
```


Has the following (22) variables: 

* `baseline value`: Baseline fetal heart rate (FHR).
* `accelerations`: Number of accelerations per second. 
* `fetal_movement`: Number of fetal movements per second. 
* `uterine_contractions`: Number of uterine contractions per second. 
* `light_decelerations`: Number of LD's per second.
* `severe_decelerations`: Number of SD's per second. 
* `prolongued_decelerations`: Number of PD's per second. 
* `abnormal_short_term_variability`: Percentage of time with abnormal short term variability. 
* `mean_value_of_short_term_variability`: Mean value of short term variability. 
* `percentage_of_time_with_abnormal_long_term_variability`: Percentage of time with abnormal long term variability.
* `mean_value_of_long_term_variability`: Mean value of long term variability. 
* `histogram_width`: Width of the histogram made using all values from a record. 
* `histogram_min`: Histogram minimum value.  
* `histogram_max`: Histogram maximum value. 
* `histogram_number_of_peaks`: Number of peaks in the exam histogram. 
* `histogram_number_of_zeroes`: Number of zeroes in the exam histogram. 
* `histogram_mode`: Hist mode. 
* `histogram_mean`: Hist mean. 
* `histogram_median`: Hist median. 
* `histogram_variance`: Hist variance. 
* `histogram_tendency`: Histogram trend. 
* `fetal_health`: Fetal health with 3 levels: 
  - 1 - Normal
  - 2 - Suspect
  - 3 - Pathological